A Software Engineering Perspective on \\ OpenMP and PThreads

### 2017-08-08 sections/00-abstract.tex
Add text for the abstract, introduction and conclusion.

\begin{abstract}
\todo[inline,color=cyan,author=Pedro]{Summarize results}
\end{abstract}

>>>>

\begin{abstract}

    Selecting the most effective tools and languages for teaching parallel and
    distributed computing presents a challenge for teachers and students. Often
    a high-level solution seems easier, but it comes with the cost of hiding
    underlying concepts and problems. We present a comparison between the
    \textit{Pthreads} and \textit{OpenMP} APIs considering the experiences and
    performance results of undergraduate and graduate Computer Science students
    from the University of São Paulo.
    We show that students were able to achieve better performance results using
    \textit{Pthreads} than \textit{OpenMP}, despite their perception that
    \textit{Pthreads} is harder to learn and use.

\end{abstract}

### 2017-08-08 sections/01-introduction.tex
Add text for the abstract, introduction and conclusion.

\todo[inline,color=cyan,author=Pedro]{Write new introduction}

...

In this work we have gathered data from assignments made by graduate students
for the \emph{Introduction to Parallel and Distributed Computing} course. Among
other exercises, the students were asked to ...

\todo[inline,color=cyan,author=Pedro]{Describe paper structure}

>>>>

Multi-core architectures became essential and ubiquitous for High-Performance
Computing in the last decade, their relevance increased by recent slowdowns on
Moore's law and the end of Dennard scaling.  Selecting the appropriate tool for
programming multi-core architectures is also essential, since it is necessary
to program a multi-core architecture effectively to leverage its performance.

Programming multi-core architectures demand different skills and tools from
what is needed for traditional sequential programming. Selecting an effective
language, framework or API for parallel and distributed programming is a
challenge for teachers and students, especially due to the quantity and
heterogeneity of available resources.

Tools with a higher level of abstraction may provide a smoother transition from
sequential to parallel and distributed programming, making it easier for
students to develop working code. The cost of high-level abstractions often is
the hiding of underlying concepts, problems and solutions, as well as code
performance.

The contribution of this paper is a comparison between \textit{Pthreads} and
\textit{OpenMP}, two commonly used APIs for parallel and distributed
programming.  The comparison considers the perceptions and performance results
from undergraduate and graduate students enrolled in the course of Concurrent,
Parallel and Distributed Computing of the University of São Paulo, Brazil.  We
show that students were able to achieve better performance results using
\textit{Pthreads} than \textit{OpenMP}, despite their perception that
\textit{Pthreads} is harder to learn and use.

The paper is organized as follows.
Section \ref{sec:apis} describes \textit{OpenMP} and \textit{Pthreads}.
Section \ref{sec:background} discusses related work.
Section \ref{sec:researchdesign} discusses the course of
Concurrent, Parallel and Distribute Computing of the University
of São Paulo, and presents the questionnaire and our research
questions.
Section \ref{sec:responses} presents, discusses and analyses
the students' responses to the questionnaire.
Section \ref{sec:conclusion} summarizes the results and
presents future work.

### 2017-08-08 sections/02-apis.tex
Add text to research design. Minor edits.
Add text for the abstract, introduction and conclusion.

In this section we briefly describe \textit{OpenMP} and \textit{Pthreads}.

@@

When \texttt{pthread\_create} is called a thread is created and the application
enters a parallel region with two simultaneous threads.  The main mechanism
available for synchronizing thread executions is the barrier. The
\textit{Pthreads} library implements barriers as calls to
\texttt{pthread\_join}. A thread that reaches a barrier will wait for other
threads. Line 10 shows an example of thread synchronization using a barrier.

>>>>

The selection of teaching tools for Parallel and Distributed courses impacts
the development of students' abilities to solve algorithmic problems
efficiently. The importance of selecting good tools for teaching Parallel and
Distributed Computing has been increasingly important, especially since the
topic became a core component of the ACM undergraduate computer science
curricula in 2013~\cite{acmcurricula}.  In this section we briefly describe
\textit{OpenMP} and \textit{Pthreads}, two commonly used tools for parallel and
distributed programming.

@@

When \texttt{pthread\_create} is called a thread is created and the application
enters a parallel region with two simultaneous threads.  The main mechanism
available for synchronizing thread executions is the barrier. The
\textit{Pthreads} library implements barriers as calls to
\texttt{pthread\_join}. A thread that reaches a barrier will wait for other
threads. Line 11 shows an example of thread synchronization using a barrier.

### 2017-08-08 sections/03-background.tex
Add text for the abstract, introduction and conclusion.

Different approaches have been used to study the selection of the best level of
abstraction for teaching Parallel and Distributed programming courses.
Falcao~\cite{6565518} argues that introducing a high-level parallel programming
interface such as \textit{OpenMP} in the beginning of an undergraduate computer
science course would cause a small overhead and benefit the rest of the course.
Foley \textit{et al.}~\cite{FOLEY2017138} support the introduction of practical
learning experiences along an undergraduate course, of increasing depth as the
course progresses. They introduce OnRamp, a web application to help students
learn to write parallel code starting with practical and moving to conceptual
experiences.  Pllana \textit{et al.}~\cite{Pllana:2009} adopt a Model Driven
Development approach to develop parallel solutions using Parallel Building
Blocks. They introduce a programming environment that transforms these
high-level models in source code. The tool's goal is to hide from the students
the complexity of writing parallel code.

@@

\todo[inline,color=orange,author=Raphael]{Maybe we should make a better
connection between the end of this section and the next one;}

>>>>

Different approaches have been proposed to study the selection of the best level of
abstraction for teaching Parallel and Distributed programming courses.
Falcao~\cite{6565518} argues that introducing a high-level parallel programming
interface such as \textit{OpenMP} in the beginning of an undergraduate computer
science course would cause a small overhead and benefit the rest of the course.
Foley \textit{et al.}~\cite{FOLEY2017138} support the introduction of practical
learning experiences along an undergraduate course, of increasing depth as the
course progresses. They introduce OnRamp, a web application to help students
learn to write parallel code starting with practical and moving to conceptual
experiences.  Pllana \textit{et al.}~\cite{Pllana:2009} adopt a Model Driven
Development approach to develop parallel solutions using Parallel Building
Blocks. They introduce a programming environment that transforms these
high-level models in source code. The tool's goal is to hide from the students
the complexity of writing parallel code.

@@

In this paper we present data that supports the importance of composing a
curriculum for Parallel and Distributed Computing courses using low-level
programming interfaces, such as \textit{Pthreads}, as well as high-level
interfaces, such as \textit{OpenMP}.  Brown \textit{et
al.}~\cite{Brown:2010:SPC:1971681.1971689}, points that the choice of a
software tool should be only one of the topics explored during a parallel
programming course. Topics such as data structures, algorithms, software design
and parallel hardware platforms should also be explored.

### 2017-08-08 sections/05-responses.tex
Add text for the abstract, introduction and conclusion.

@@

\begin{figure}[htpb]
    \centering
    \includegraphics[width=0.85\columnwidth]{classes_questions}
    \caption{Student API knowledge \textit{before} the assignment and their relation to classes}
    \label{fig:classes}
\end{figure}

@@

@@

\begin{figure}[htpb]
    \centering
    \includegraphics[width=0.85\columnwidth]{likert_questions}
    \caption{Student responses to the Likert Scale questions in Table \ref{tab:likert}}
    \label{fig:likert}
\end{figure}

@@

\begin{figure}[htpb]
    \centering
    \includegraphics[width=0.85\columnwidth]{comparisons}
    \caption{Student responses to the questions in Table \ref{tab:comparisons}}
    \label{fig:comparisons}
\end{figure}

\todo[inline,color=cyan,author=Pedro]{Improve these last paragraphs}

The responses presented in this section enable the conclusion that the students
perceived less difficulty using and learning \textit{OpenMP} in the context
of the assignment. They also reported the perception that using \textit{OpenMP}
made it easier to improve the performance of the sequential code we provided.
Our analysis of the submissions revealed that the majority of the students
($52.2\%$) achieved better performance in the assignment with either
\textit{OpenMP} ($17.4\%$) or \textit{Pthreads} ($34.8\%$). Contradicting the
perception of the students, the implementation using \textit{Pthreads} achieved
better performance in twice the submissions.

This shows that the perceptions of usability and performance improvement
regarding \textit{OpenMP} did not reflect in the actual implementations and
submissions. We assert that \textit{OpenMP} is still not as easy as it appears
to be, and that the students' responses strengthen the argument for teaching
lower-level technologies for parallel and distributed computing.

>>>>

\begin{figure}[htpb]
    \centering
    \includegraphics[width=0.85\columnwidth]{classes_questions}
    \caption{Students' responses to questions in Table \ref{tab:classes}}
    \label{fig:classes}
\end{figure}

@@

@@

\begin{figure}[htpb]
    \centering
    \includegraphics[width=0.85\columnwidth]{likert_questions}
    \caption{Students' responses to the Likert Scale questions in Table \ref{tab:likert}}
    \label{fig:likert}
\end{figure}

@@

@@

\begin{figure}[htpb]
    \centering
    \includegraphics[width=0.85\columnwidth]{comparisons}
    \caption{Students' responses to the questions in Table \ref{tab:comparisons}}
    \label{fig:comparisons}
\end{figure}
